EAST: {\underline E}xpression Fragment {\underline A}ssembly from {\underline S}panning {\underline T}rees

### 2011-06-09 papers/EAST.bioinf/east-gb.tex
papers/EAST.bioinf: New version of the EAST paper, re-structured for the journal Bioinformatics.  Semi-final draft for submission.

\author{Yuan Zhang$^1$
  \email{Yuan Zhang - zhangy9@muohio.edu}
  \and
  Dhananjai M Rao$^1$
  \email{Dhananjai Rao - raodm@muohio.edu}
  \and
  Mufit Ozden$^1$
  \email{Mufit Ozden - ozdenm@muohio.edu}
  \and
  Jens Mueller$^2$
  \email{Jens Mueller - muellej@muohio.edu}
  \and
  Chun Liang\correspondingauthor$^{1,3}$
  \email{Chun Liang\correspondingauthor - liangc@muohio.edu}
  and
  John E. Karro\correspondingauthor$^{1,4.5}$
  \email{John E. Karro\correspondingauthor - karroje@muohio.edu}
}

@@

\address{
    \iid(1)Department of Computer Science and Software Engineering,  \\
    \iid(2)Informational Technology Services
    \iid(3)Department of Botanty, \\
    \iid(4)Department of Microbiology, \\
    \iid(5)Department of Statistics, Miami University, Oxford OHIO, USA
}

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
  \section*{Introduction} {\color{blue} The {\it de novo} assembly of
    RNA transcripts is an important and computationally challenging
    problem, made more difficult by the advent of Next Generation
    Sequencing technologies such as 454 and Illumina
    \cite{Nagaraj07,Rao10}.  Upon completion of
    transcriptome-sequencing, the investigator is left with a large
    set of sequence fragments but no ordering or transcript
    association information.  These fragments must be clustered by
    transcript and assembled before the sequencing output can be truly
    useful.  {\it de novo} assembly tools assemble the
    transcript out of the fragments without using genmic sequence
    information.  {\it de novo} tools are obivously necessary when working with a genome that is
    unsequenced or poorly annotated, but is also useful when there is
    limited knowledge of gene splice sites, as well as for the discovery
    of novel events \cite{Birol09,Robertson10}.}

@@

\textcolor{blue}{In assessing result quality, we must view the \peace\/ clustering tool
\cite{Rao10} and \east\/ as a combined process.  While most assembly
tools integrate these two jobs, \east\/ assumes clustering has been
performed and requires the \peace\/ MST output.  Hence for purposes of
the following section, we consider them as a single tool \peast, and
assess the quality of these results.}

@@

\subsection*{Measuring Quality}
Our primary measure of quality was the {\bf normalized a-score}, based
on the {\it assembly score} (a-score) used in {\it Liang et al.}
\cite{Liang00}.  For an assembly of some original sequence, the
a-score is defined as:
$$\mbox{($2\times$sequence length) - ($5\times$number of
mismatches) - ($15 \times$number of indels)}$$
with a ``perfect'' score equal to twice the sequence length.  Our
normalized a-score modifies this definition as:
$$\frac{\mbox{($2\times$assembly length) - ($5\times$number of mismatches)
    - ($15\times$number of indels)}}{\mbox{($2\times$sequence
    length)}}$$ with a perfect score equal to 1.  This normalization
allows for the comparison of assembly quality between sequences of
different lengths. Further, using this score we can more accurately
quantify the quality of a partial reconstruction: if we are able to
reconstruct only a portion of a fragment, the normalized a-score will
take full consideration of the quality of the reconstruction, but then
bound the score by the relative size of the reconstruction.
\textcolor{blue}{(For example, were a tool only able to reconstruct
  one fourth of a transcript -- but able to reconstruct that fraction
  perfectly -- the resulting contig's a-score would be $0.25$.)}  In
addition we look at: the number of contigs produced by a tool
normalized by the number of transcripts (a value equal to 1 in a
perfect reconstruction); the number of singleton ESTs in the solution
(i.e. ESTs unassigned to any contig and thus, most cases, effectively
thrown away), and the runtime of the tool.

@@

\noindent {\bf Sanger Sequencing:} \textcolor{blue}{We first compare the
  assembly quality of the tools on a Sanger Sequencing model, using
  \east, \capthree, and \tgicl\/, the tools designed to work with
  Sanger sequence.  The quality of any assembly will be directly
  affected by the {\it covreage} of the input: the average number of
  ESTs that cover a based.  Thus we investigate the a-score as a
  function of error rate at different coverage levels.}  For each
parameter combination we applied each tool to 30 independent, randomly
generated simulated sets, derived from 15 randomly chosen members of
our zebra fish gene set, reporting the average normalized a-score.  In
Figure~\ref{sangerAscore} we look at normalized a-score as a function
of error rate, fixing coverage at 50, 30 and 10 and varying base read
error rate from 0\% to 6\% for each of the three tools designed for
Sanger Sequencing.  While all three tools perform well when applied to
error free (i.e. impossible) sequencing technology, we find \east\/ to
be considerably more robust to base-call errors than the other tools.
Taking the mid-range coverage value of 30 as an example, at a 1\%
error rate we see \east, \capthree, and \tgicl\/ all achieving an
essentially perfect normalized a-score.  At a 3\% error rate \east\/
maintains its almost perfect normalized a-score, showing a 2\%
improvement over \tgicl\/ and a 21\% improvement over \capthree.  At a
4\% error rate \east\/ still maintains its near-perfect score (showing
a 54\% and 85\% improvement over \capthree\/ and \tgicl\/
respectively).  Even at a 6\% error rate \east\/ is still achieving a
normalized a-score of 0.95 (a 110\% improvement over the next closest
tool, \tgicl\/), meaning that at most 5\% of the bases were left out
of their contigs or were incorrectly reconstructed.

>>>>

\author{Yuan Zhang$^1$
  \email{Yuan Zhang - zhangy9@muohio.edu}
  \and
  Dhananjai M Rao$^1$
  \email{Dhananjai Rao - raodm@muohio.edu}
  \and
  Mufit Ozden$^1$
  \email{Mufit Ozden - ozdenm@muohio.edu}
  \and
  Jens Mueller$^2$
  \email{Jens Mueller - muellej@muohio.edu}
  \and
  Praveen Kumar Raj Kumar$^4$
  \email{Praveen Kumar - rajkump@muohio.edu}
  \and
  Chun Liang\correspondingauthor$^{1,3}$
  \email{Chun Liang\correspondingauthor - liangc@muohio.edu}
  and
  John E. Karro\correspondingauthor$^{1,4,5}$
  \email{John E. Karro\correspondingauthor - karroje@muohio.edu}
}

@@

\address{
  \iid(1)Department of Computer Science and Software Engineering,  \\
  \iid(2)Informational Technology Services\\
  \iid(3)Department of Botanty, \\
  \iid(4)Department of Microbiology, \\
  \iid(5)Department of Statistics, Miami University, Oxford OHIO, USA
}

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
  \section*{Introduction}

  The {\it de novo} assembly of RNA transcripts is an important and
  computationally challenging problem, made more difficult by the
  advent of Next Generation Sequencing technologies such as 454 and
  Illumina \cite{Nagaraj07,Rao10}.  Upon completion of
  transcriptome-sequencing, the investigator is left with a large set
  of sequence fragments but no ordering or transcript association
  information.  These fragments must be clustered by transcript and
  assembled before the sequencing output can be truly useful.  {\it de
    novo} assembly tools assemble the transcript out of the fragments
  without using genomic sequence information.  {\it de novo} tools are
  obivously necessary when working with a genome that is unsequenced
  or poorly annotated, but are also useful when there is limited
  knowledge of gene splice sites \cite{Birol09,Robertson10}, they can
  aid in the discovery of novel biological processes
  (e.g. transplicing and fusion transcript) \cite{Mitelman07,Li10b}, and
    they are invaluable for annotating structural variations not
    contained within the genome \cite{Li08}.

@@

In assessing result quality, we must view the \peace\/ clustering tool
\cite{Rao10} and \east\/ as a combined process.  While most assembly
tools integrate these two jobs, \east\/ assumes clustering has been
performed and requires the \peace\/ MST output.  Hence for purposes of
the following section, we consider them as a single tool \peast, and
assess the quality of these results.

@@

\subsection*{Measuring Quality}
Our primary measure of quality was the {\bf normalized a-score}, based
on the {\it assembly score} (a-score) used in {\it Liang et al.}
\cite{Liang00}.  For an assembly of some original sequence, the
a-score is defined as:
$$\mbox{($2\times$sequence length) - ($5\times$number of
mismatches) - ($15 \times$number of indels)}$$
with a ``perfect'' score equal to twice the sequence length.  Our
normalized a-score modifies this definition as:
$$\frac{\mbox{($2\times$assembly length) - ($5\times$number of mismatches)
    - ($15\times$number of indels)}}{\mbox{($2\times$sequence
    length)}}$$ with a perfect score equal to 1.  This normalization
allows for the comparison of assembly quality between sequences of
different lengths. Further, using this score we can more accurately
quantify the quality of a partial reconstruction: if we are able to
reconstruct only a portion of a fragment, the normalized a-score will
take full consideration of the quality of the reconstruction, but then
bound the score by the relative size of the reconstruction.
(For example, were a tool only able to reconstruct a
  contig covering one fourth of a transcript -- but able to
  reconstruct that fraction perfectly -- the resulting contig's
  normalized a-score would be $0.25$.)  In addition we look at: the
number of contigs produced by a tool normalized by the number of
transcripts (a value equal to 1 in a perfect reconstruction); the
number of singleton ESTs in the solution (i.e. ESTs unassigned to any
contig and thus, most cases, effectively thrown away), and the runtime
of the tool.

@@

\noindent {\bf Sanger Sequencing:} We first compare the
  assembly quality of the tools on a Sanger Sequencing model, using
  \east, \capthree, and \tgicl\/, the tools designed to work with
  Sanger sequence.  The quality of any assembly will be directly
  affected by the {\it covreage} of the input: the average number of
  ESTs that cover a based.  Thus we investigate the a-score as a
  function of error rate at different coverage levels.  For each
parameter combination we applied each tool to 30 independent, randomly
generated simulated sets, derived from 15 randomly chosen members of
our zebra fish gene set, reporting the average normalized a-score.  In
Figure~\ref{sangerAscore} we look at normalized a-score as a function
of error rate, fixing coverage at 50, 30 and 10 and varying base read
error rate from 0\% to 6\% for each of the three tools designed for
Sanger Sequencing.  While all three tools perform well when applied to
error free (i.e. impossible) sequencing technology, we find \east\/ to
be considerably more robust to base-call errors than the other tools.
Taking the mid-range coverage value of 30 as an example, at a 1\%
error rate we see \east, \capthree, and \tgicl\/ all achieving an
essentially perfect normalized a-score.  At a 3\% error rate \east\/
maintains its almost perfect normalized a-score, showing a 2\%
improvement over \tgicl\/ and a 21\% improvement over \capthree.  At a
4\% error rate \east\/ still maintains its near-perfect score (showing
a 54\% and 85\% improvement over \capthree\/ and \tgicl\/
respectively).  Even at a 6\% error rate \east\/ is still achieving a
normalized a-score of 0.95 (a 110\% improvement over the next closest
tool, \tgicl\/), meaning that at most 5\% of the bases were left out
of their contigs or were incorrectly reconstructed.
